Negative Selection in the 20th Century

### 2020-05-21 why-negative-selection.Rmd
Reorganize paper. Add intro and appendix. Weighted regressions for parents. Correlations beween effect sizes and correlations with income/education.

```md
# TODO

* Why so many missings for age_fulltime_edu?
  - ah, need f.6138 for those who went to college
  - but note that YearsEdu.ISCED has many fewer missings
  - ask Abdel how it was calculated
  - use this for survey weighting?

* Look at geography, esp. in siblings regressions.

* Work more on weighting the data?
  - weight by age at first live birth
  - education using ISCED?
  - geography?
  - ask for existing work by others?

* control for age in n_partners regressions

## Data to gather
* f.2139 - age first had sex (includes "never had sex" which may explain
  some of the many NAs for f.2141, num sex partners)

@@

options(digits = 2)
knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = FALSE, 
        error = TRUE
      )

@@

@@

@@

@@

drake::loadd(famhist)
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
          seq(40, 65, 5),
          extend = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
      pivot_longer(
        -YOB, 
        names_to = "score_name",
        values_to = "score"
      )

```{r calc-means-over-time-children, eval = FALSE}
pgs_over_time_children <- famhist %>% 
      mutate(
        YOB_kids = (year_flb + year_llb)/2,
        YOB      = santoku::chop(
                     YOB_kids, 
                     brk_left(seq(1960, 1995, 5), close_end = FALSE), 
                     seq(60, 90, 5),
                     extend = FALSE
                   )
      ) %>% 
      filter(! is.na(YOB)) %>% 
      group_by(YOB) %>% 
      summarize(across(all_of(score_names), 
        ~weighted.mean(.x, n_children, na.rm = TRUE)
      )) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score")

pgs_over_time <- bind_rows(
        Respondents = pgs_over_time,
        Children    = pgs_over_time_children,
        .id         = "Generation"
      )

@@

We run regressions on two dependent variables:

@@

We run regressions both with and without controls for the 100 top principal
components of the genetic data.

```{r regs-data}

@@

res_all_wide <- res_all %>% 
  select(reg.type, dep.var, term, estimate) %>% 
  pivot_wider(names_from = c(reg.type, dep.var), values_from = estimate) %>% 
  mutate(
    consistent = sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
      sign(`raw_N siblings`) == sign(`raw_N children`) &
      sign(`raw_N siblings`) == sign(`controlled_N children`)
  )

n_pgs  <- as.double(nrow(res_all_wide))
prop_consistent <- mean(res_all_wide$consistent)
controlled_smaller_sibs <- abs(res_all_wide[["controlled_N siblings"]]) < 
  abs(res_all_wide[["raw_N siblings"]])
controlled_smaller_children <- abs(res_all_wide[["controlled_N children"]]) < 
  abs(res_all_wide[["raw_N children"]])
median_prop_sibs <- median(abs(res_all_wide[["controlled_N siblings"]]) /
    abs(res_all_wide[["raw_N siblings"]]))
median_prop_children <- median(abs(res_all_wide[["controlled_N children"]]) /
    abs(res_all_wide[["raw_N children"]]))

prop_gens_raw <- res_all_wide[["raw_N children"]]/res_all_wide[["raw_N siblings"]]
prop_gens_controlled <- res_all_wide[["controlled_N children"]]/res_all_wide[["controlled_N siblings"]]
median_prop_gens_raw <- median(prop_gens_raw[prop_gens_raw > 0])
median_prop_gens_controlled <- median(prop_gens_controlled[prop_gens_controlled > 0])
```

@@

Estimates are broadly consistent across generations. For 
`r sum(res_all_wide$consistent)` out of `r n_pgs` polygenic scores, all 4 
estimates have the same sign.

However, effect sizes are much smaller for *children* than *siblings*
regressions. Among consistently-signed estimates, the median effect size for
children as a proportion of the effect size for  siblings is 
`r median_prop_gens_raw`, or `r median_prop_gens_controlled` with controls.

In siblings regressions, effect sizes are smaller when controlling for
principal components -- sometimes much smaller, as in the case of height. 
`r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled" effect sizes have 
a smaller absolute value than the corresponding "raw" effect size. The median
proportion between raw and controlled effect sizes is `r median_prop_sibs`.
Among the children regressions, this no longer holds. Effect sizes are barely
affected by controlling for principal components.

```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children.", fig.height = 7}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      labs(x = "Effect size", y = "") +
      facet_wrap(vars(dep.var), ncol = 2) +
      pval_scale() +
      my_fill_scale(n = 2, guide = guide_legend(title = "")) +
      theme(
        legend.position = "bottom", 
        legend.margin   = margin(2, 2, 2, 2)
      )

@@

@@

# Selection over time

```{r period-regs-data}

@@

Negative selection seems to decrease over time. Figure \@ref(fig:plot-by-period) shows
effect sizes for *number of siblings* and *number of children*, median-split by
parents' year of birth and own year of birth respectively.  Parents' year of
birth is imputed, which is likely to produce some bias.

```{r plot-by-period, fig.cap = "Effect sizes of PGS on number of children/siblings by own/parents' year of birth. PGS with the largest mean effect sizes are shown.", fig.height = 7, fig.width = 6}

max_abs <- function (x) max(abs(x))
res_period %>% 
      mutate(
        dep.var       = ifelse(children, "Children", "Siblings"),
        period        = (period == "late") + children,
        score_name    = pretty_names(score_name),
        score_name    = fct_reorder(score_name, estimate, max_abs, 
                          .desc = TRUE),
        p             = diff.p.value
      ) %>% 
      arrange(score_name) %>% 
      slice(1:48) %>% 
ggplot(aes(period, estimate, colour = dep.var, group = dep.var)) + 
      geom_hline(yintercept = 0, colour = "grey30") + 
      geom_point(size = 1.5) +
      geom_line(aes(linetype = p < 0.05/{{n_tests}})) +
      facet_wrap(vars(score_name), nrow = 4) + 
      pval_scale() +
      my_fill_scale(n = 2, aesthetics = "colour") +
      scale_linetype_manual(values = c("FALSE" = "dashed", "TRUE" = "solid"))

@@

By definition, the sibling regressions exclude members of the parents'
generation who had no children. This is likely to bias results towards zero, since
much of the effect in children regressions is due to respondents with high scores
being more likely to have no children. So, we cannot directly compare 
effect sizes for the two sets of regressions. Within the sibling regressions,
the most common pattern is that negative effects shrink in absolute size 
(Table \@ref(tab:siblings-by-period)).

```{r siblings-by-period}

      add_footnote(glue("Significance is measured at p < 0.05/{n_tests}"), 
        border = 0.4) %>% 
      set_contents(1, everywhere, 
        c("Change", "Number of scores")) %>% 
      set_right_border(final(1), 1, 0) %>% 
      set_top_padding(1) %>% 
      set_bottom_padding(1) %>%
      set_left_padding(everywhere, 1, 0) %>% 
      set_right_padding(everywhere, 2, 0) %>% 
      set_caption(caption) %>% 
      set_caption_width(1)
}

@@

In children regressions, no clear pattern is visible 
(Table \@ref(tab:children-by-period)).

```{r children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in effect sizes between early and late born respondents, 'children' regressions")
```

```{r res-weighted}

mean_prop_wtd_unwtd <- res_weighted_all %>% 
      filter(sign(estimate_weighted) == sign(estimate_unweighted)) %>% 
      summarize(mean = mean(prop_wtd_unwtd)) %>% 
      pull(mean)

@@

Effect sizes tend to be smaller for children regressions. This
could be caused by ascertainment bias in the UK Biobank sample -- e.g.,
if respondents themselves are a more selected sample than the respondents' 
parents. To check this, we weighted UK Biobank participants by age of leaving
full time education. Our sample weights are based on the 2006 General Household
Survey, calculating proportions within age and sex cells among White British
respondents. We then rerun the basic *children* regressions.

Figure \@ref(fig:plot-weighted) shows the results, with unweighted estimates
plotted for comparison. Weighting increases effect sizes on average by 
`r 100 * (mean_prop_wtd_unwtd - 1)` per cent. These weights are only basic 
attempts to correct for ascertainment bias. Effect sizes might be increased 
further by more precise weighting.

@@

res_weighted_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate_weighted, order_abs(1))
      ) %>% 
      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
        names_to = "Regression", values_to = "estimate") %>% 
      mutate(
        Regression = sub("estimate_", "", Regression)
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = Regression,
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      labs(x = "Effect size", y = "") +
      my_fill_scale(n = 2) +
      pval_scale(aesthetics = "colour")

@@

# Causality

```{r pgs-corrs}
cor_tbl <- cor(famhist[score_names], use = "pair") %>% 
      as_tibble() %>% 
      mutate(score_name = score_names, .before = 1) %>% 
      pivot_longer(cols = -score_name, values_to = "Correlation") %>% 
      filter(score_name < name) %>% 
      arrange(desc(Correlation)) %>% 
      slice(1:10) %>% 
      mutate(across(c(score_name, name), pretty_names))

as_hux(cor_tbl, add_colnames = TRUE) %>% 
      set_contents(1, 1:2, c("PGS", "PGS")) %>% 
      theme_basic(header_col = FALSE) %>% 
      set_bottom_border(1, everywhere, 0.4) %>% 
      set_bottom_border(final(1), everywhere, 0.4) %>% 
      set_caption("Top 10 correlations between polygenic scores")

@@

Different polygenic scores are correlated. Table \@ref(tab:pgs-corrs) shows the top
correlations in the sample. Because of this, bivariate correlations between
PGS and number of children might be driven by other genetic scores. To explore which polygenic scores are driving negative selection, we run a single
omnibus regression of *number of children* on all the PGS. We exclude EA2, 
waist-hip ratio, waist-circumference, and "Hip combined" since they are 
highly correlated with other scores, which could make our estimates unstable. Figure \@ref(fig:plot-together) shows the results. Interestingly, several PGS remain independently significant, although
effect sizes are reduced.

We next examine how different subgroups contribute to natural selection.

## Males and females

```{r}

drake::loadd(res_sex)

female_lower <- res_sex %>% 
      pivot_wider(id_cols = c(term), names_from = sex, 
        values_from = c(estimate, conf.low, conf.high)) %>% 
      mutate(female_lower = estimate_Female < estimate_Male) %>% 
      pull(female_lower)
```

Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
separately for males and females. For `r sum(female_lower)` out of
`r length(female_lower)` PGS, selection is more negative for women
than for men. Differences are particularly large for educational attainment and height PGS.

n_regs <- as.double(nrow(res_sex))
res_sex %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2, direction = -1) +
        my_labs()

```{r corr-age-flb}

@@

Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
women only,
controlling for age at first live birth. Effect sizes are greatly reduced.
In `r opp_signed` out of `r length(score_names)` cases, they are of the 
opposite sign. The correlation between effect sizes
controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.

```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}

@@

res_combined %>% 
      mutate(
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        pval_geom_point() +
        pval_scale() +
        scale_fill_brewer(type = "qual", palette = 3) +
        my_labs()
```

@@

res_age_flb_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(estimate, term, colour = p.value < 0.05 / {{n_regs}})) + 
      my_vline +
      pval_geom_point(fill = "steelblue4") +
      pval_scale() + 
      my_labs()

@@

## Number of sexual partners

Figure \@ref(fig:plot-n-partners) splits males and females up by lifetime number
of sexual partners. Remarkably, across both sexes, negative selection is
strongly reversed for respondents who had 3 or fewer sexual partners in their
lifetime.

```{r plot-n-partners, fig.cap = "Effect sizes on number of children by number of sexual partners", fig.height = 7, fig.width = 6}

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = `N partners`, 
        colour = p.value < 0.05/{{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex)) +
        my_fill_scale(n = 2) +
        pval_scale() +
        theme(legend.position = "bottom") +
        my_labs(x = "", y = "")

@@

## Education and income

Figure \@ref(fig:plot-educ-levels) splits respondents up by education levels.
Both negative and positive selection are typically larger and more significant
for those who left school before 16. Table \@ref(tab:educ-levels) summarizes
the results.

>>>>

@@

options(digits = 2)
options(dplyr.summarise.inform = FALSE) # shut up about summarize
knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = FALSE, 
        error = TRUE,
        fig.height = 3.5
      )

@@

my_theme <- function (...) theme(axis.text.y = element_text(size = 7), ...)

@@

# Introduction

Existing work suggests that natural selection is taking place in modern populations,
but that effect sizes are small. In this paper we investigate natural selection
in the UK over the twentieth century, using data from UK Biobank. We use
questions on number of children and number of siblings to reconstruct two
generations of family size per respondent. We then relate these to polygenic
scores for several different characteristics.

Natural selection is taking place on most polygenic scores, and is broadly
"negative" in that characteristics that most people would consider undesirable
are being selected for. Effect sizes are larger for number of siblings than for
number of children. This is not because the strength of natural selection has
changed over time: effect sizes do not differ between early- and late-born
respondents, within either children or siblings. Instead, the difference is
probably due to ascertainment bias in UK Biobank. Indeed, natural selection is
concentrated among population subgroups that are under-represented in the
sample, including less-educated people and poorer households. When we weight 
the sample to correct for ascertainment bias, many effect sizes are considerably 
larger.

Among mothers who were older at first birth, and among parents with fewer sexual
partners, effect signs are reversed so that natural selection is broadly
"positive". The effects of PGS can be decomposed into two channels: an effect on
age at first live birth, and an opposite-signed effect on number of children
conditional on age at first live birth. This pattern may be explained by
economic theories of fertility. In these, higher potential earnings have two
opposite effects on fertility: a fertility-increasing income effect (higher
income makes children more affordable), and a fertility-lowering substitution
effect (time spent on childrearing has a higher cost in foregone earnings).
Among couples who can take advantage of coordination in childcare activities,
the income effect dominates; among single parents or those in unstable
relationships, the substitution effect dominates.

# Open questions/TODO

* If this theory is true then the pattern of differences is explained by
  (potential) earnings. So the strength of the relationship of a PGS with 
  number of children should be explained by the strength of its relationship
  with earnings and education.
* Even after weighting, effect sizes for siblings are still higher than for
  children. Why?
  - Remaining ascertainment bias?
  - Siblings don't include those with 0 children?
  - Something else?
* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Could be useful: f.2139 - age first had sex (includes "never had sex" 
  which may explain some of the many NAs for f.2141, num sex partners)
*  Need f.6138 to correctly calculate `age_fulltime_edu` including
  those who went to college.
* Look at geography, esp. in siblings regressions.
* Work more on weighting the data
  - check for existing work by others?
  - deal with the fact that GHS 06 has its own weights and use
    subset.svydesign to reweight?
* control for age in n_partners regressions

@@

Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores by 5-year
birth intervals, for the entire sample. Several scores show consistent increases
or declines over this 30-year period, of the order of 5% of a standard deviation.

@@

drake::loadd(famhist)
drake::loadd(ghs_weights)
famhist <- left_join(famhist, ghs_weights, by = "f.eid")
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
          seq(40, 65, 5),
          extend = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score")

@@

We run bivariate regressions on two dependent variables:

@@

```{r calc-regs-all}

@@

res_all_wide <- res_all %>% 
  filter(reg.type == "raw") %>% 
  select(dep.var, term, estimate) %>% 
  pivot_wider(names_from = dep.var, values_from = estimate)

n_pgs  <- as.double(nrow(res_all_wide))

prop_gens <- res_all_wide[["N siblings"]]/res_all_wide[["N children"]]
median_prop_gens <- median(prop_gens[prop_gens > 0])
```

@@

Estimates are broadly consistent across generations. However, effect sizes are 
much larger for *siblings* than for *children* 
regressions. Among consistently-signed estimates, the effect size for
siblings is `r 100*median_prop_gens` per cent at the median.

```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children."}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      filter(reg.type == "raw") %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = dep.var, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      my_labs() +
      pval_scale() +
      my_fill_scale(n = 2, direction = -1, guide = guide_legend(title = "")) +
      my_theme()

@@

We next examine how different subgroups of the population contribute to natural 
selection. These analyses have two goals: to get a sense of how ascertainment bias
could affect our basic results, and to learn more about the causes of natural 
selection in the UK population. Because we have more data on respondents than on
their parents, we focus on *children* regressions.

## Sex, income and education

Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
separately for males and females. Differences are particularly large for
educational attainment, height and MDD. There is no overall pattern in
differences of effect size, however.

drake::loadd(res_sex)
n_regs <- as.double(nrow(res_sex))

res_sex %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2, direction = -1) +
        my_labs() +
        my_theme()

@@

Figure \@ref(fig:plot-educ-level) splits respondents up by education levels.
Selection effects are typically larger and more significant for those who left
school before 16.  Figure \@ref(fig:plot-income) splits respondents by household
income category. A very similar pattern holds, with selection effects being
larger for those in the poorest income category.

```{r plot-educ-level, fig.cap = "Effect sizes on number of children by age left full-time education"}

drake::loadd(res_edu)
n_regs <- as.double(nrow(res_edu))
res_edu %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
      ) %>% 
      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
        fill = `Age left FTE`)) +
        my_vline +
        pval_geom_point() + 
        pval_scale() +
        my_fill_scale(n = 3, direction = -1) +
        my_labs() +
        my_theme()

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        )
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = Income, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point(alpha = 0.8) + 
        my_fill_scale(n = 5, direction = -1) +
        pval_scale() +
        my_labs() +
        my_theme()
```

```{r calc-income-edu-controlled}
drake::loadd(res_income_controlled, res_edu_controlled)
n_regs_ic <- nrow(res_income_controlled)
n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)

The UK Biobank sample overrepresents highly educated people and high-income
households. This suggests that statistics from the sample may underestimate
the level of natural selection in the population.

```{r calc-age-flb-cross}

drake::loadd(res_age_flb_cross)

cor_age_flb <- res_age_flb_cross %>% 
      mutate(category = gsub(":.*", "", term)) %>% 
      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}

@@

Age at first live birth correlates with number of children ever born. Figure
\@ref(fig:plot-age-flb-cross) shows *children* regressions estimated separately
for each tercile of age at first live birth. Several effects are strikingly
different across terciles. ADHD and MDD are selected for amongst the youngest
third of mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several PGS of body
measurements are selected against only among older mothers. The correlation
between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.
This is the first sign that selection may work in different directions across
different subgroups.

Figure \@ref(fig:plot-n-partners) splits males and females by lifetime number of
sexual partners, at the median value of 3. Again, selection effects are reversed
across the groups. Among men and women with more than 3 sexual partners, for
example, EA is selected against; among those with 3 or fewer sexual partners, it
is selected for. Selection is broadly negative (positive) among those 
with more (less) partners.

```{r plot-age-flb-cross, fig.cap = "Effect sizes on number of children, by age at first live birth terciles"}

drake::loadd(res_age_flb_cross)

# this counts each cross term as a separate test:
n_regs <- as.double(nrow(res_age_flb_cross))

res_age_flb_cross %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1)),
        term       = gsub("age_flb_cat(.*):.*", "\\1", term)
      ) %>% 
      ggplot(aes(estimate, score_name, fill = term, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 3, direction = -1, 
          guide = guide_legend(title = "Age at FLB")) +
        my_labs() +
        my_theme()

@@

```{r plot-n-partners, fig.cap = "Effect sizes on number of children by lnumber of sexual partners", fig.height = 4}

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = `N partners`, 
        colour = p.value < 0.05/{{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex)) +
        my_fill_scale(n = 2) +
        pval_scale() +
        my_theme(legend.position = "bottom") +
        my_labs(x = "", y = "")

@@

```{r calc-res-weighted}

mean_prop_wtd_unwtd <- mean(res_weighted_all$prop_wtd_unwtd)

As a fix for ascertainment bias in the UK Biobank sample, we weight participants
by sex, age (within the 40-71 age group), and education. Our sample weights are
calibrated based on the 2006 General Household Survey. We then rerun the basic
*children* regressions. Figure \@ref(fig:plot-weighted) shows the results, with
unweighted estimates plotted for comparison. Weighting increases effect sizes on
average by `r 100 * (mean_prop_wtd_unwtd - 1)` per cent.

@@

res_weighted_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate_weighted, order_abs(1))
      ) %>% 
      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
        names_to = "Regression", values_to = "estimate") %>% 
      mutate(
        Regression = sub("estimate_", "", Regression)
      ) %>% 
      ggplot(aes(y = term, x = estimate, shape = Regression,
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      geom_point(size = 1.5, stroke = 0.7, fill = "steelblue") +
      labs(x = "Effect size", y = "") +
      scale_shape_manual(values = c("weighted" = 22, "unweighted" = 21)) +
      pval_scale(aesthetics = "colour") +
      my_theme()

@@

```{r plot-sibs-weighted, eval = FALSE}

# what this shows is that even after weighting, sibling effect sizes are
# still stronger. In fact weighting hasn't changed things much.
drake::loadd(res_weighted, res_sibs_weighted)
n_regs <- as.double(nrow(res_weighted))
# bind_rows(children = res_weighted, sibs = res_sibs_weighted, .id = "Dep. Var.") %>% 
res_all %>% 
      filter(reg.type == "raw") %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = `dep.var`,
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      labs(x = "Effect size", y = "") +
      my_fill_scale(n = 2) +
      pval_scale(aesthetics = "colour")

@@

```{r calc-flb-weights}

drake::loadd(res_flb_weights)
drake::loadd(res_sex)
res_flb_all <- left_join(
        res_flb_weights, 
        res_sex %>% filter(sex=="Female") %>% select(-subset, -sex), 
        by = "term", 
        suffix = c(".wt", ".nowt")
      ) %>% 
      mutate(prop_estimate = estimate.wt/estimate.nowt)

prop_wt_nowt <- mean(res_flb_all$prop_estimate)
pct_change_score <- function (score) {
  100 * (res_flb_all$prop_estimate[res_flb_all$term == score] - 1)
}
```

We also weighted data for women with children only, by age, education and age 
at first live birth.  Figure \@ref(fig:plot-flb-weights) 
shows the results. Weighting increased effect sizes on average by 
`r 100 * (prop_wt_nowt - 1)` per cent. There 
are large changes for some variables, including EA3 
(`r pct_change_score("EA3_excl_23andMe_UK")` per cent), cognitive
ability (`r pct_change_score("cognitive_ability")` per cent) and
waist-hip ratio (WHR; `r pct_change_score("whr_combined")` percent).

```{r plot-flb-weights, fig.cap = "Effect sizes on number of children, female respondents weighted by age at first live birth"}

bind_rows(
        Weighted   = res_flb_weights, 
        Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, -sex),
        .id        = "Regression"
      ) %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(estimate, term, shape = Regression, 
        color = p.value < 0.05/66)) +
        my_vline +
        geom_point(size = 1.5, stroke = 0.7, fill = "steelblue") +
        scale_shape_manual(values = c("Weighted" = 22, "Unweighted" = 21)) +  
        pval_scale() +
        my_labs() + 
        my_theme()
```

These results suggest that previous work may have underestimated the level of
selection occurring in the population. [XXX who?]

```{r calc-corr-age-flb}

@@

## Mechanisms behind natural selection

In this section, we dig into the mechanisms causing natural selection in the
population. In particular, why are selection effects *oppositely* signed in
certain subgroups -- for example, in older and younger mothers (by age at first
live birth) or in those with more and fewer lifetime sexual partners?

A possible answer is given by the economic theory of fertility [XXX cite Becker 1960].
According to this, increases in potential earned income affect fertility via two opposing channels. There is an "income effect" by which children become more
affordable, like any other good. Since childrearing has a cost in time, there is
also a substitution effect: the opportunity cost of childrearing increases if
one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children.

Traditionally, it was assumed that since women
spent more of their time on childcare, the substitution effect would dominate
for women's earnings, while the income effect would dominate for men's earnings. 
However, a more relevant split may be between those raising children alone, and
those raising children in a two-parent household. The substitution effect is likely
to be particularly strong for lone parents, since they have less opportunity
to share childcare responsibilities.

If so, then genetic characteristics which affect one's earnings potential in the
labour market may lead to opposing effects on fertility. Genetic
variants which improve one's earnings may increase fertility among couples, but
decrease fertility among single parents.

This explanation predicts that the strength of polygenic scores' effects on
fertility will correlate with the strength of their relationship with earnings.
We measure the correlation of each PGS with household income, as well as with
education level, a phenotypic variable which is likely to predict earnings. We
then correlate these measurements with effect sizes on number of children.
Across scores, there is a clear negative correlation between effect on
fertility, and link to both income and education.

```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children and correlations with income and education"}

drake::loadd(res_all)
cors_income <- cor(famhist[c(score_names)], famhist$income_cat, use = "pairwise")
# TODO: use age_left_edu when this includes college
cors_educ <-   cor(famhist[c(score_names)], famhist$edu_qual, use = "pairwise")

effect_size <- res_all %>% 
      filter(dep.var == "N children", reg.type == "raw") %>% 
      pull(estimate)

dfr <- data.frame(
      type = rep(c("Cor. with income", "Cor. with educ"), each = 33),
      cor = c(cors_income[,1], cors_educ[,1]),
      effect_size = rep(effect_size, 2)
)

ggplot(dfr, aes(effect_size, cor, colour = type)) + 
      geom_point() +
      facet_wrap(vars(type)) + 
      my_vline + 
      my_hline +
      labs(x = "Effect on n children", y = "Correlation") +
      scale_colour_manual(values = c("darkblue", "darkgreen")) +
      theme(legend.position = "none")

# Bits to integrate

Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
women only,
controlling for age at first live birth. Effect sizes are greatly reduced. Partly,
this is because the regressions exclude childless women. However,
in `r opp_signed` out of `r length(score_names)` cases, effect sizes actually
change sign with the controls. The correlation between effect sizes
controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.
```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}

@@

res_combined %>% 
      mutate(
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        pval_geom_point() +
        pval_scale() +
        scale_fill_brewer(type = "qual", palette = 3) +
        my_labs() +
        my_theme()
```

@@

res_age_flb_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(estimate, term, colour = p.value < 0.05 / {{n_regs}})) + 
      my_vline +
      pval_geom_point(fill = "steelblue4") +
      pval_scale() + 
      my_labs() +
      my_theme()

@@

# Appendix

## Controlling for principal components

Polygenic scores could capture effects that are really due to population 
stratification, although would not change our results for natural selection of
the scores. In \@ref(fig:plot-regs-controlled) we show results for selection on
polygenic scores residualized for the top 100 principal components of the
genetic data, calculated within the UK Biobank population. (XXX Abdel details.)

```{r calc-regs-controlled}

res_all_controlled <- res_all %>% 
  select(reg.type, dep.var, term, estimate) %>% 
  pivot_wider(names_from = c(reg.type, dep.var), values_from = estimate) %>% 
  mutate(
    consistent = sign(`raw_N siblings`) == sign(`raw_N children`) &
                 sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
                 sign(`raw_N siblings`) == sign(`controlled_N children`)
  )

n_pgs  <- as.double(nrow(res_all_controlled))
prop_consistent_controlled <- mean(res_all_controlled$consistent)
controlled_smaller_sibs <- abs(res_all_controlled[["controlled_N siblings"]]) < 
  abs(res_all_controlled[["raw_N siblings"]])
controlled_smaller_children <- abs(res_all_controlled[["controlled_N children"]]) < 
  abs(res_all_controlled[["raw_N children"]])
median_prop_sibs <- median(abs(res_all_controlled[["controlled_N siblings"]]) /
    abs(res_all_controlled[["raw_N siblings"]]))
median_prop_children <- median(abs(res_all_controlled[["controlled_N children"]]) /
    abs(res_all_controlled[["raw_N children"]]))

@@

In siblings regressions, effect sizes are smaller when residualizing for
principal components -- sometimes much smaller, as in the case of height. 
`r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled" effect sizes have 
a smaller absolute value than the corresponding "raw" effect size. The median
proportion between raw and controlled effect sizes is `r median_prop_sibs`.
Among the children regressions, this no longer holds. Effect sizes are barely
affected by controlling for principal components.

Overall, `r prop_consistent_controlled * 100` per cent of effect sizes are
consistently signed across all four regressions (on children and siblings, and
with and without residualization).
